Output Summary

### 2022-04-21 03-post-process/output-plots.Rmd
Update all residual diagnostic text

# Residual Diagnostics {.tabset .tabset-pills}

@@

@@

This section is devoted to evaluating how consistent the estimated noise terms are with the assumptions of the random processes they are assumed to be generated by.
Specifically, the noise (or "residual") terms should be:

* **Mean-Zero**: each year has an expected value for the realized outcome, and across years errors should cancel out such that the expected residual in any given year is zero.
* **IID**: all residuals are generated by the same random process, i.e., with mean zero and SD equal to one; with the exception of cases with an auto-regressive process included residuals among years are assumed independent.
* **Lack of time series patterns** is implied by these two assumptions, i.e., there should be no time trends or cyclical "regime shifts".

The presence of egregious violations in these assumptions is an indication that the model may give predictions that are inconsistent with the historical data and assumed hypotheses/mechanisms.

@@

To examine whether these assumptions were consistently met or violated, residuals from the process model and observation model were calculated and standardized.

Standardized residuals were calculated differently depending on the type of random process:

* **Log-normal** (e.g., for abundance data likelihoods): `(log(realized) - log(expected))/SD`, i.e., noise in log space
* **Logit-normal** (e.g., for survival data likelihoods, survival process noise): `(logit(realized) - logit(expected))/SD`, i.e., noise in logit space
* **Binomial and multinomial** (e.g., for carcass count likelihoods): `sqrt(realized) - sqrt(expected)`; for multinomial counts, these were calculated for each category separately, i.e., the Freeman-Tukey statistic, which is similar to the $\chi^2$ statistic but better handles cases where some bin probabilities are small.

For observation model residuals, `SD` represents the assumed observation error variability supplied to the model as known. For process model residuals, `SD` represents the model-estimated `sig` parameter associated with that random process.

This standardization enables comparison across model components that are on different scales (e.g., abundance vs. probability terms) and have different variances, either process or observation.
The specific standardization used implies that the mean of the residuals from a given model component should be zero across all years, should have a standard deviation of 1, and 95% of the residuals should fall between approximately -2 and 2.

@@

For each random process in the model, the time series of standardized noise terms is plotted below (mean and 95% CRI in red, first and last half mean values in blue). To summarize these time series quantitatively, three across-year summary statistics are reported:

* **Across-year mean** -- intended to verify that noise is centered on the expected value overall (ideally equal to 0)
* **Across-year SD** -- intended to verify that noise is as variable as expected overall (ideally equal to 1)
* **mean(last half) - mean(first half)** -- intended to verify that noise is not time-trending (ideally equal to 0)

@@

The posterior mean of the mean noise value across all years; bold indicates the 95% CRI did not encompass 0.

Values:

* Close to 0 indicate that the time series of noise terms was not biased relative to the expected value (i.e., assumption met)
* Below 0 indicate that the time series of noise terms was biased low relative to the expected value
* Above 0 indicate that the time series of noise terms was biased high relative the expected value

@@

The posterior mean of the SD of noise value across all years; bold indicates the 95% CRI did not encompass 1 (all noise terms were standardized).

Values:

* Close to 1 indicate that the time series of noise terms was as variable as expected (i.e., assumption met)
* Below 1 indicate that the time series of noise terms was less variable than expected
* Above 1 indicate that the time series of noise terms was more variable than expected

@@

Values:

* Close to 0 indicate a lack of an overall time trend (i.e., assumption met)
* Below 0 indicate an overall declining time trend in noise terms
* Above 0 indicate an overall increasing time trend in noise terms

>>>>

# Random Noise Diagnostics {.tabset .tabset-pills}

@@

This section is devoted to evaluating how consistent the estimated noise terms are with the assumptions of the random processes they are assumed to be generated by.
Diagnostics are separate by process model and observation model and the time series of noise terms is displayed to visually detect issues.
Each model component (process or observation) includes summary tables that quantitatively summarize the results displayed on the time series figures.

@@

The presence of egregious assumption violations is an indication that the model may give predictions that are inconsistent with the historical data and assumed hypotheses/mechanisms.
If the assumptions are reasonably met, the noise (or "residual", representing inter-annual variability) terms should be:

1. **Non-existent on average**: each year has an expected value and the noise term represents how far the realized value falls from the expectation, either higher or lower. Thus, the expectation for any given noise term is "no noise"; i.e., if the same year was replicated many times over under the same random process, the average of the realized values would be the expected value.
2. **IID**: all noise terms for a process are generated by the same random process, i.e., with mean zero and common variance; noise terms are assumed independent from year to year, except in cases with an auto-regressive process included.
3. **Free of time series patterns**: this is implied by the above two assumptions, i.e., there should be no time trends or cyclical "regime shifts."

@@

Because the model used several different types of probability distributions for accounting for process and observational noise (logit-normal, log-normal, binomial, and multinomial), we wished to use a standardization technique that would enable using the same metrics to assess assumption validity for all random processes.
We chose to use a quantile standardization, wherein the noise term for a given process in a given year is expressed as a value between zero and one and is obtained as the cumulative probability of the realized value when treating the posterior predictive distribution for that year as the CDF.

For example, consider an arbitrary outcome $x_y$ in a given year from a random process with density/mass function $f()$ with parameter vector $\mathbf{\theta}$ and lower bound of $l$ (e.g., $l=-\infty$ or $l = 0$, depending on the domain of $f()$).
The standardized residual is then:

$$
\Pr(X \leq x_y) = \int_{l}^{x_y} f(v; \mathbf{\theta}) dv
$$

If repeated realizations $x_y, x_{y+1},\dots,x_{y+n}$ are consistent with the random process $f$, then their quantile-standardized values will be distributed uniformly between zero and one (i.e., $\Pr(X \leq x_y) \sim \mathrm{U}(0,1)$).
If the parameter vector $\mathbf{\theta}$ varies annually, then $\mathbf{\theta}_y$ can be substituted and the same uniform property should hold if the realized outcomes are consistent with both (_a_) the random process $f()$ and (_b_) the (probably deterministic) process generating inter-annual variability in $\mathbf{\theta}_y$.

The interpretation of these quantile-standardized noise terms is:

* **Equal to 0.5**: The realized outcome was exactly in the center of $f()$
* **Less than 0.5**: The realized outcome was somewhere in the lower tail of $f()$, values closer to zero are more extreme noise outcomes
* **Greater than 0.5**: The realized outcome was somewhere in the upper tail of $f()$, values closer to one are more extreme noise outcomes

Because the parameter vector $\theta$ has a posterior distribution, we calculated these standardized noise terms for each MCMC sample, which provides posterior distributions of the noise terms as well.

We performed this calculation for each random process (of both the observation and process components) in the model using the built-in cumulative distribution functions in JAGS.
E.g., for some arbitrary survival term `phi` which has logit-normal process noise around the expected value `mu_phi` with logit-scale standard deviation of `sigma_phi`:

```
# priors
mu_phi ~ dbeta(1,1)
sigma_phi ~ dunif(0,5)

for (y in 1:ny) {
  # random process
  Lphi[y] ~ dnorm(logit(mu_phi), 1/sigma_phi^2)
  phi[y] <- ilogit(Lphi[y])

  # standardized noise term
  phi_qresid[y] <- pnorm(logit(phi[y]), logit(mu_phi), 1/sigma_phi^2)
}
```

Continuing with this example, suppose we have counts of tagged fish we know survived (`x_surv[y]`) and the total number of fish tagged (`n_tagged[y]`). We can calculate the noise term of `x_surv[y]` when assuming it is a binomial random variable using:

```
for (y in 1:ny) {
  # likelihood statement for fitting model
  x_surv[y] ~ dbin(phi[y], n_tagged[y])

  # standardized noise term
  x_surv_qresid[y] <- pbin(x_surv[y], phi[y], n_tagged[y])
}
```

For multinomial random variables (i.e., adult age/origin composition), we used category-specific binomial distributions to assess noise.
This allowed assessing the validity of age/origin-specific noise terms needed to accommodate the data.

@@

For each random process in the model, the time series of standardized noise terms is plotted below:

* red points/bars: year-specific posterior mean/95% CRI
* blue horizontal lines: first and last half of time series mean values
* grey horizontal lines: expected 25%, 50%, and 75% quantiles of the standardized residuals).

To summarize these time series quantitatively, we report three across-year summary statistics:

* **Across-year mean** -- intended to verify that noise terms are centered on the expected value overall (ideally equal to 0.5; the expected value of a $\mathrm{U}(0,1)$ distribution)
* **Across-year SD** --- intended to verify that noise terms are as variable as expected overall (ideally equal to $\sqrt{1/12}$ = `r round(sqrt(1/12), 2)`; the standard deviation of a $\mathrm{U}(0,1)$ distribution)
* **mean(last half) - mean(first half)** -- intended to verify that noise is not time-trending (ideally equal to 0)

@@

The posterior mean of the mean standardized noise value across all years; bold indicates the 95% CRI did not encompass the target value of 0.5.

**Interpretation of Values**:

* **Close to 0.5** indicate that the time series of noise terms was not biased relative to the expected value (i.e., assumption met)
* **Below 0.5** indicate that the time series of noise terms was biased low relative to the center of the distribution
* **Above 0.5** indicate that the time series of noise terms was biased high relative the center of the distribution

@@

The posterior mean of the SD of noise value across all years; bold indicates the 95% CRI did not encompass $\sqrt{1/12}$ = `r round(sqrt(1/12), 2)`; the standard deviation of a $\mathrm{U}(0,1)$ distribution.

**Interpretation of Values**:

* **Close to `r round(sqrt(1/12), 2)`** indicate that the time series of noise terms was as variable as expected (i.e., assumption met)
* **Below `r round(sqrt(1/12), 2)`** indicate that the time series of noise terms was less variable than expected
* **Above `r round(sqrt(1/12), 2)`** indicate that the time series of noise terms was more variable than expected

@@

**Interpretation of Values**:

* **Close to 0** indicate a lack of an overall time trend (i.e., assumption met)
* **Below 0** indicate an overall declining time trend in noise terms
* **Above 0** indicate an overall increasing time trend in noise terms
